---
title: "Lab 05 - Tempe Accident Analysis"
author: "Your Name"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    theme: flatly
    highlight: haddock
    toc: yes
    toc_float: yes
---

```{r include = FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10)     # Knitting options

library(dplyr)
library(pander)                           # Load required packages

```

<br>

# Source Data

The following analysis uses data comprised of traffic accidents in the city of 
Tempe, Arizona. They constitute 28,470 observations and 32 variables. You can 
find more details in the dataset documentation at the 
[**Tempe Open Data Portal**](https://data.tempe.gov/dataset/high-severity-traffic-crashes-1-08). 

```{r include = FALSE}

# READ IN DATA

url <- paste0("https://github.com/DS4PS/Data-Science-Class/blob",
              "/master/DATA/TempeTrafficAccidents.rds?raw=true")

dat <- readRDS(gzcon(url(url)))     # Method per instructions

```

<br>

## Preview

We can explore the first six observations in the following table using function `head()`.

```{r}

head(dat)

```

<br>
<br>

# Preprocessing

All data preprocessing are available in **Appendix A: Data Preprocessing**.

```{r include = FALSE}

date.vec <- strptime(dat$DateTime, 
                     format = "%m/%d/%y %H:%M")       # Create date fields

dat$hour   <- format(date.vec, format = "%H")
dat$month  <- format(date.vec, format = "%b")
dat$day    <- format(date.vec, format = "%a")
dat$day365 <- format(date.vec, format = "%j")
dat$week   <- format(date.vec, format = "%V")

dat$day <- factor(dat$day, 
                  levels = c("Mon",
                             "Tue",
                             "Wed",
                             "Thu",
                             "Fri",
                             "Sat",
                             "Sun"))                  # Order weekdays

dat$hour12 <- format(date.vec, 
                     format="%l %p")                  # Create 12-hour format

time.levels <- c("12 AM", paste(1:11, "AM"), 
                 "12 PM", paste(1:11, "PM"))

dat$hour12 <- factor(dat$hour12, 
                     labels = time.levels)            # Order time intervals

age.labels <- paste0("Age ", 
                     c(16,18,25,35,45,55,65,75), "-", 
                     c(18,25,35,45,55,65,75,100) )

dat$age <- cut(dat$Age_Drv1, 
               breaks = c(16,18,25,
                          35,45,55,
                          65,75,100), 
               labels = age.labels)                   # Discretize age ranges

```

<br>
<br>

# Part I: Summaries

The following provides brief exploratory summaries on key data points.

<br>

## 1. Mondays Accidents

**Question:** *How many accidents happened on Mondays?*

**Note:** *This solution has been completed for you.*

<br>

```{r}

sum(dat$day == "Mon", 
    na.rm = TRUE)

```

<br>

**Answer:** **4,094** accidents occurred on Mondays.

<br>

## 2. Monday Accidents (%)

**Question:** *What proportion of accidents each week occur on Monday??*

<br>

```{r}

# Code

```

<br>

**Answer:** **[X]%** of all accidents occur on Mondays.

<br>

## 3. Harmful Monday Accidents (%)

**Question:** *What proportion of accidents on Mondays result in harm??*

**Note:** *"Harm" is defined as any accident that causes at least one injury or* 
*fatality, a.k.a. a "casualty".*

<br>

```{r}

# Code

```

<br>

**Answer:** **[X]%** of all Monday accidents have at least one casualty.

<br>

## 4. Most Common Accidents

**Question:** *What is the most common accident type on Mondays?*

<br>

```{r}

dat %>%
  filter(day == "Mon") %>%
  count(Collisionmanner) %>%
  arrange(desc(n))

```

<br>

**Answer:** **[X]** collisions are the most common accident types.

<br>

## 5. Differences in Accidents

**Question:** *Are there differences in the proportion of accidents that result in harm each day of the week?*

<br>

```{r}

# Code

```

<br>

**Answer:** Significantly more accidents occur **[X]**.

<br>

## 6. Accident Stats by Weekday

**Instructions:** Create a table that reports the following for each day of the week.

* Total accidents
* Total injuries
* Total fatalities
* Proportion of harmful accidents (i.e. casualties)

<br>

```{r}

dat %>%
  group_by(day) %>%
  mutate(harm_acc = Totalinjuries > 0 | Totalfatalities > 0) %>%
  summarize(n = n(),
            injuries = sum(Totalinjuries),
            fatalities = sum(Totalfatalities),
            harm.rate = mean(harm_acc))

```

<br>
<br>

# Part II: Age Groups

The following provides summaries of accidents by age groups.

<br>

## 1. Accidents by Hour & Age

**Instructions:** Create a table of counts of accidents by time of day (`hour12`) and age of driver (`age`).

**Question:** *Which age group has the largest number of accidents at 7 AM?*

<br>

```{r}

# Code

```

<br>

**Answer:** Drivers of **[X]** have the greatest number of accidents from 7:00 to 7:59 AM.

<br>

## 2. Accidents by Hour & Age (%)

**Instructions:** Create a new table of time of day and age group that reports the proportion of accidents at "7 AM" (`hour12`) for each age group (`age`). *The proportions within each age group should sum to one.*

<br>
<br>

**Objective:** The table should contain the following columns and initial values:

```{r echo = FALSE}

data.frame(age = c("Age 16-18", "Age 18-25", "Age 25-35"),
           hour12 = c("7 AM", "7 AM", "7 AM"),
           n = c(77, 408, 371),
           n.age = c(1458, 8796, 5456),
           n.hour = c(1606, 1606, 1606),
           p = c(0.05, 0.25, 0.23),
           p.age = c(0.05, 0.05, 0.07),
           p.hour = c(0.05, 0.25, 0.23)) %>%
  pander()

```

<br>
<br>

**Solution:** *The solution has been partially provided. Use the hints provided to create the table.*

```{r}

dat %>% 
  group_by(hour12, age) %>% 
  summarize(n = n()) %>%      # Partial solution
  group_by(age) %>%           # Requires two new variables in 'mutate()'
  mutate()                    # Lastly, requires function 'filter()

```

<br>
<br>

# Part III: Rates of Harm

The following reports the accidents, casualties, proportion, and average 
casualties per harmful accident.

<br>

## 1. Accidents by Hour

**Instructions:** Visualize total accidents by time of day (`hour`).

**Note:** *This solution has been completed for you.*

<br>

```{r}

dat %>%
  group_by(hour) %>%
  summarize(n = n()) %>%
  plot(type = "b",
       bty = "n",
       pch = 19,
       cex = 2,
       xlab = "Hour",
       ylab = "Total Accidents",
       main = "Total Accidents by Time of Day")

```

<br>

## 2. Total Casualties by Hour

**Instructions:** Visualize total injuries and fatalities by time of day (`hour`).

<br>

```{r}

# Code

```

<br>

## 3. Accidents with Casualties (%)

**Instructions:** Visualize the proportion of harmful collisions our of all collisions by time of day (`hour`).

<br>

```{r}

# Code

```

<br>

## 4. Avg. Harm by Harmful Accident

**Instructions:** Visualize average injuries or fatalities per harmful collision.

<br>

```{r}

# Code

```

<br>

# Appendix

The following code is used for preprocessing tasks in the above solutions.

<br>

## A: Data Preprocessing

These expressions transform, format, and create myriad new variables used in this report.

<br>

```{r eval = FALSE, echo = TRUE, include = TRUE}

date.vec <- strptime(dat$DateTime, 
                     format = "%m/%d/%y %H:%M")       # Create date fields

dat$hour   <- format(date.vec, format = "%H")
dat$month  <- format(date.vec, format = "%b")
dat$day    <- format(date.vec, format = "%a")
dat$day365 <- format(date.vec, format = "%j")
dat$week   <- format(date.vec, format = "%V")

dat$day <- factor(dat$day, 
                  levels = c("Mon",
                             "Tue",
                             "Wed",
                             "Thu",
                             "Fri",
                             "Sat",
                             "Sun"))                  # Order weekdays

dat$hour12 <- format(date.vec, 
                     format="%l %p")                  # Create 12-hour format

time.levels <- c("12 AM", paste(1:11, "AM"), 
                 "12 PM", paste(1:11, "PM"))

dat$hour12 <- factor(dat$hour12, 
                     levels = time.levels)            # Order time intervals

age.labels <- paste0("Age ", 
                     c(16,18,25,35,45,55,65,75), "-", 
                     c(18,25,35,45,55,65,75,100) )

dat$age <- cut(dat$Age_Drv1, 
               breaks = c(16,18,25,
                          35,45,55,
                          65,75,100), 
               labels = age.labels)                   # Discretize age ranges

```

<br>
<br>

<center>------------<h4> **DELETE THIS LINE & ALL LINES BELOW BEFORE SUBMITTING** </h4>------------</center>

<br>
<br>

#### **How to Submit**

Use the following instructions to submit your assignment, which may vary depending on your course's platform.

<br>
<br>

#### **Knitting to HTML**

When you have completed your assignment, click the "Knit" button to render your `.RMD` file into a `.HTML` report.

<br>
<br>

#### **Special Instructions**

Perform the following depending on your course's platform:

* **Canvas:** Upload both your `.RMD` and `.HTML` files to the appropriate link
* **Blackboard or iCollege:** Compress your `.RMD` and `.HTML` files in a `.ZIP` file and upload to the appropriate link

`.HTML` files are preferred but not allowed by all platforms.

<br>
<br>

#### **Before You Submit**

Remember to ensure the following before submitting your assignment.

1. Name your files using this format: **Lab-##-LastName.rmd** and **Lab-##-LastName.html**
2. Show both the solution for your code and write out your answers in the body text 
3. Do not show excessive output; truncate your output, e.g. with function `head()`
4. Follow appropriate styling conventions, e.g. spaces after commas, etc.
5. Above all, ensure that your conventions are consistent

See [Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml) for examples of common conventions.

<br>
<br>

#### **Common Knitting Issues**

`.RMD` files are knit into `.HTML` and other formats procedural, or line-by-line.

* An error in code when knitting will halt the process; error messages will tell you the specific line with the error
* Certain functions like `install.packages()` or `setwd()` are bound to cause errors in knitting 
* Altering a dataset or variable in one chunk will affect their use in all later chunks
* If an object is "not found", make sure it was created or loaded with `library()` in a previous chunk

**If All Else Fails:** If you cannot determine and fix the errors in a code chunk that's preventing you from knitting your document, add `eval = FALSE` inside the brackets of `{r}` at the beginning of a chunk to ensure that R does not attempt to evaluate it, that is: `{r eval = FALSE}`. This will prevent an erroneous chunk of code from halting the knitting process.  

<br>
<br>
<br>