---
title: "Lab 06 - Dataset Joins"
author: "Your Name"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    theme: flatly
    highlight: haddock
    toc: yes
    toc_float: yes
---

```{r include = FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10)     # Global knitr options

library(dplyr)
library(pander)
library(Lahman)                           # Load packages

data(Teams)
data(Salaries)                            # Load data

options(scipen = 999)                     # Disable sci. notation

```

<br>

# Introduction

The following work identifies the baseball teams that have achieved the largest 
number of wins on the smallest budgets since 1985 per the `Lahman` data package.

<br>

# Preprocessing

This analysis links two `Lahman` package tables, viz. `Salaries` and `Teams`, 
to determine total wins per season and the salary cost of each win earned. See 
the following preview of the first observations in `Salaries`:

<br>

```{r echo = FALSE}

head(Salaries) %>%
  pander()

```

<br>

See the following preview of select variables and first observations in `Teams`:

<br>

```{r echo = FALSE}

Teams %>%
  select(yearID, teamID, lgID, G, W, L) %>%
  head() %>%
  pander()

```

<br>

Lastly, `Salaries` data are adjusted for inflation at a constant annual increase 
rate of 3%. Observe:

<br>

```{r echo = TRUE}

Salaries <- Salaries %>% 
  mutate(salary.adj = salary * (1.03)^(max(yearID) - yearID))

head(Salaries) %>%
  pander()

```

<br>

In order to calculate the cost per win, adjusted individual player salaries from 
the `Salaries` table must be aggregated by team and year in the `Teams` table.

<br>
<br>

# Step 1: Aggregating Salaries

Using the `Salaries` table and grouping on variables `yearID` and `teamID` via
package `dplyr` function `group_by()`, the following uses function 
`summarize()` to *calculate and name* the following summary variables for 
*each year and team*: 

<br>

* Total adjusted team budget using `salary.adj` and `sum()`
 
* Total players using `n()`

<br>

```{r}

# Your code here

```

<br>
<br>

# Step 2: Join Salaries & Teams

The following merges the `Salaries` summary data and full `Teams` tables using 
either base R function `merge()` or package `dplyr` functions `left_join()`. 

<br>

* Variables `teamID` and `yearID` are unique merging keys
 
* Example: `dataset_1 %>% left_join(dataset_2)`

<br>

```{r}

# Your code here

```

<br>
<br>

# Step 3: Calculate Cost per Win

The following uses the merged `Salaries` summary data and `Teams` data to 
create a new variable with function `mutate()` that calculates the total 
team salary cost per win, divided by $100,000.

<br>

* Divide (`/`) total salaries from **Step 1** by total wins (`W`)
 
* Divide (`/`) that value by `100000`

<br>

```{r}

# Your code here

```

<br>
<br>

# Step 4: Select, Filter, & Arrange

The following filters out teams in merged data with less than 25 players on 
their annual roster. It then arranges teams in ascending order by cost per win. 
Lastly, only variables are preserved to create the full version of the 
table preview.

<br>

* Use `filter()` to remove observations with less than 25 players
 
* Use `arrange()` to order observations by total cost per win
 
* Use `select()` to preserve only essential variables

<br>

**Objective:** The following contains variables and initial values 
of the full table.  Variable names may differ.

<br>

```{r echo = FALSE}

data.frame(yearID = c(1988, 1986, 1987),
           teamID = c("PIT", "TEX", "ML4"),
           lgID = c("NL", "AL", "AL"),
           Rank = c(2, 2, 3),
           G = c(160, 162, 162),
           W = c(85, 87, 91),
           n = c(27, 26, 28),
           team.budget = c(13724134, 16367320, 17186960),
           cost.per.win = c(1.614604, 1.881301, 1.888677)) %>%
  pander()

```

<br>

**Solution:** Create the full version of the table below.

```{r}

# Your code here

```

<br>
<br>

<center>------------<h4> **DELETE THIS LINE & ALL LINES BELOW BEFORE SUBMITTING** </h4>------------</center>

<br>
<br>

#### **How to Submit**

Use the following instructions to submit your assignment, which may vary depending on your course's platform.

<br>
<br>

#### **Knitting to HTML**

When you have completed your assignment, click the "Knit" button to render your `.RMD` file into a `.HTML` report.

<br>
<br>

#### **Special Instructions**

Perform the following depending on your course's platform:

* **Canvas:** Upload both your `.RMD` and `.HTML` files to the appropriate link
* **Blackboard or iCollege:** Compress your `.RMD` and `.HTML` files in a `.ZIP` file and upload to the appropriate link

`.HTML` files are preferred but not allowed by all platforms.

<br>
<br>

#### **Before You Submit**

Remember to ensure the following before submitting your assignment.

1. Name your files using this format: **Lab-##-LastName.rmd** and **Lab-##-LastName.html**
2. Show both the solution for your code and write out your answers in the body text 
3. Do not show excessive output; truncate your output, e.g. with function `head()`
4. Follow appropriate styling conventions, e.g. spaces after commas, etc.
5. Above all, ensure that your conventions are consistent

See [Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml) for examples of common conventions.

<br>
<br>

#### **Common Knitting Issues**

`.RMD` files are knit into `.HTML` and other formats procedural, or line-by-line.

* An error in code when knitting will halt the process; error messages will tell you the specific line with the error
* Certain functions like `install.packages()` or `setwd()` are bound to cause errors in knitting 
* Altering a dataset or variable in one chunk will affect their use in all later chunks
* If an object is "not found", make sure it was created or loaded with `library()` in a previous chunk

**If All Else Fails:** If you cannot determine and fix the errors in a code chunk that's preventing you from knitting your document, add `eval = FALSE` inside the brackets of `{r}` at the beginning of a chunk to ensure that R does not attempt to evaluate it, that is: `{r eval = FALSE}`. This will prevent an erroneous chunk of code from halting the knitting process.  

<br>
<br>
<br>
